\documentclass[11pt,a4paper]{article}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage{csquotes}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{isabelle,isabellesym}

\providecommand{\nexists}{\not\exists}

\usepackage[top=3cm,bottom=4.5cm]{geometry}

% this should be the last package used
\usepackage{pdfsetup}

\urlstyle{rm}
\isabellestyle{tt}

\title{Byzantine-tolerant detection of causality:\\
       there is no holy grail}
\author{Mechanisation in Isabelle/HOL of the impossibility results of\\
        Misra and Kshemkalyani (Parallel Computing 124, 2025)}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
We mechanise in Isabelle/HOL the impossibility results of
Misra and Kshemkalyani, \emph{Byzantine-tolerant detection of
causality: there is no holy grail} (Parallel Computing 124 (2025),
article 103136).  The causality determination problem
$\mathrm{CD}(E,F,e_i^*)$ asks an asynchronous distributed algorithm
to reconstruct the global execution history $E$ at a correct
process $p_i$, with respect to a target event $e_i^*$, such that
the happened-before relation evaluated against the algorithm's
collected $F$ agrees with the truth in $E$.  In the presence of
even a single Byzantine process this is impossible.

We formalise:

\begin{itemize}
  \item Theorems 1 and 2 (paper Section 4.1): no algorithm can
        prevent false negatives in the Byzantine model, and for
        internal events no algorithm can prevent both false
        negatives and false positives.  Both proofs are
        constructive: given any candidate algorithm we exhibit a
        Byzantine adversary using a fresh natural number as event
        identifier.
  \item Theorems 3, 4, 5 (paper Section 4.2): the headline
        impossibility results for unicast, broadcast, and
        multicast communication.  The proofs route directly
        through Theorem 1 (\verb|CD_FN_unavoidable|) -- bypassing
        the paper's $\mathrm{Consensus} \preceq \mathrm{Black\_Box}
        \preceq \mathrm{CD} + \mathrm{FLP}$ chain at the
        critical-path level.  The paper's chain is preserved as a
        fully-proven alternative derivation
        (\verb|Reductions.thy| + \verb|BlackBox_Unsolvable.thy| +
        \verb|FLP_Consensus.thy|), with the FLP impossibility
        imported from the AFP entry \emph{FLP}
        (Bisping et al., 2025) as a proven theorem
        (\verb|flp_consensus_unsolvable|, discharged against
        \verb|flpPseudoConsensus.ConsensusFails|).
  \item Theorems 6, 7, 8 (paper Section 4.3): the
        Byzantine-happened-before extension CD\_B.  T6/T7
        solvable under unicast/broadcast at three layers (abstract
        \verb|naive_cd_B_alg| under \verb|correct_reporting|;
        operational reduction to a structural delivery property;
        concrete operational construction via an inductive
        \verb|run_step| with explicit in-flight buffer).  T8
        impossible under multicast.
  \item Theorems 9 through 14 (paper Section 4.4 + 4.5):
        cryptography variants.  Impossibilities (T9, T10, T11)
        as direct corollaries of T3/T4/T5 -- the Theorem-1 FN
        attack is crypto-independent.  Possibilities (T12, T13)
        as direct corollaries of T6/T7.  T14 (the genuinely new
        content: multicast \verb|+| crypto possible, whereas T8
        said impossible without crypto) via the naive algorithm
        under \verb|correct_reporting|.  Cryptography is treated
        at the same abstraction the paper itself takes when
        citing Bracha 1987 for BRB.
  \item Theorems 15 and 16 (paper Section 5.1): in the Byzantine
        model CD is harder than Consensus, directly from CD
        unsolvability and the abstract Consensus witness; in the
        crash-failure model, the asymmetry reverses (Consensus is
        unsolvable by FLP, CD is solvable via the
        \verb|cd_alg_with_recv| signature with $F := \mathit{recv}$
        whenever reports are faithful to the true execution).
  \item Theorems 17 and 18 (paper Section 5.2): CD and CO are
        interreducible in the Byzantine model (T17), and CO is
        subject to FN and FN-or-FP for internal-event witnesses
        (T18).  CO is mechanised as a strict restriction of CD to
        receive-event targets; T17 forward direction is
        constructive (a CD solver is automatically a CO solver),
        T18 is proven directly via a two-message fresh-id
        construction, and the reverse direction of T17 is proven
        under Byzantine premises by routing through the two
        independently-established impossibility chains.
\end{itemize}

The development sits on top of the AFP entry \emph{FLP}.  The
headline impossibility theorems (3, 4, 5) take exactly one mild
finiteness side hypothesis \verb|fin_cd| --- shape-compatible
with the \verb|fin_F| of Theorem 1 (\verb|CD_FN_unavoidable|) ---
and no other meta-level assumptions.  No HOL axioms, no locale
axioms on the impossibility chain.

A scheduler-level refinement \verb|Causal_Scheduler.thy|
operationally realises Byzantine Causal Broadcast's causal-order
delivery property: \verb|causal_run_step| is a strict refinement
of \verb|run_step| with freshness at send and a
Byzantine-happened-before predecessor causal precondition at
receive; the resulting Byzantine-happened-before version of BCB's
causal-order property
(\verb|causal_run_satisfies_bhb_causal_order|) is what paper
Definition 3 calls for, and end-to-end composition into T7
(\verb|fair_drained_causal_run_solves_CD_B_broadcast|) closes the
chain with no operational hypothesis left to the user.

The proof routes directly through Theorem 1: \verb|CD_solvable|
extracts a \verb|produces_valid_F| witness, which Theorem 1
defeats with an admissible Byzantine adversary.  The paper's
``Consensus $\preceq$ Black\_Box $\preceq$ CD + FLP'' chain is
preserved as paper-faithful documentation (R1, R2 in
\verb|Reductions.thy|; the BlackBox impossibility
\verb|BlackBox_unsolvable| in \verb|BlackBox_Unsolvable.thy|; and
the FLP impossibility \verb|flp_consensus_unsolvable| in
\verb|FLP_Consensus.thy|, proven against the AFP entry's
\verb|ConsensusFails|) but is not on the critical path of the
headline theorems.

\medskip

\noindent\emph{License:} BSD-3-Clause (matching the AFP \emph{FLP}
entry).  The source paper is open access under CC-BY 4.0 and
redistributed in this repository under that license.
\end{abstract}

\tableofcontents

\newpage

\section*{Reading order}

The theories are intended to be read in the following order:

\begin{enumerate}
  \item \verb|ByzantineSystem|: the process partition (correct
        $\uplus$ Byzantine) and the abstract Consensus signature.
  \item \verb|Events|: events, per-process and global histories,
        program order, message order, happened-before relation
        (paper Definitions 1, 2).
  \item \verb|CD|: the Causality Determination problem,
        $\mathit{valid}(F)$, false positives / negatives, the
        adversary model, the CD-solver signature (paper
        Definition 5).
  \item \verb|BlackBox|: the Black\_Box problem of paper Section
        4.2 with its piecewise broadcast value $w$.
  \item \verb|Reductions|: the two reductions of paper Section
        4.2 -- R1 (Consensus $\preceq$ BB) constructive in
        declarative Isar, R2 (BB $\preceq$ CD) constructive modulo
        the paper's named meta-level step
        \verb|cd_can_identify_correct|.
  \item \verb|Theorems_1_2|: paper Section 4.1, Theorems 1 and 2,
        with explicit Byzantine adversary constructions using
        fresh natural numbers.
  \item \verb|BlackBox_Unsolvable|: a proof of
        $\lnot$\,\verb|BlackBox_solvable procs correct| by direct
        reduction to Theorem 1 via the BB-to-CD projection.
        Discharges what was previously a meta-level hypothesis on
        Theorems 3/4/5.
  \item \verb|FLP_Consensus|: the FLP-style consensus predicate
        with the impossibility proven against the AFP entry's
        \verb|ConsensusFails|.  Retained as the AFP-FLP citation
        motivating the paper's chosen chain; not on the
        impossibility path of this development.
  \item \verb|Impossibility|: paper Section 4.2, Theorems 3, 4,
        5 -- the headline results.
  \item \verb|Foundation_Vacuity|: a regression diagnostic
        documenting why the abstract \verb|solves_Consensus|
        predicate is too weak to express FLP, and motivating the
        FLP-style chain used by \verb|Impossibility|.
  \item \verb|CD_vs_Consensus|: paper Section 5.1, Theorems 15
        and 16 (Byzantine: CD harder than Consensus; crash:
        Consensus half of T16 only).
  \item \verb|BHB|, \verb|CD_B_Algorithm|, \verb|Delivery|,
        \verb|Execution_Model|: paper Section 4.3.  The
        B-happened-before relation; the CD\_B problem; abstract
        and operational solvability under unicast/broadcast
        (Theorems 6, 7); impossibility under multicast (Theorem
        8); inductive execution model with deadlock freedom.
  \item \verb|Liveness|: real-world fairness on the inductive
        execution model.  Infinite executions modelled as
        \verb|nat => 'p config| with a side condition that each
        adjacent pair satisfies \verb|run_step|; the fairness
        predicate \verb|fair_run| states that every buffered
        triple eventually leaves the buffer; the headline
        theorem \verb|fair_run_delivers| says every correct-to-
        correct \verb|Send| in a fair infinite run has a matching
        \verb|Receive| at some later step.
  \item \verb|Primitives|: Byzantine Reliable Unicast (BRU) and
        Byzantine Causal Broadcast over Byzantine Reliable
        Broadcast (BCB-over-BRB) named explicitly at the event
        level.  \verb|bru_satisfied|, \verb|bcb_causal_order|,
        and \verb|bcb_over_brb_satisfied| predicates; operational
        discharge of BRU from the run model; end-to-end
        composition theorems giving operational T6 / T7 directly
        from a fair drained run.
  \item \verb|T6_Concrete|: a fully-concrete worked example of
        T6.  A two-process scenario with an explicit three-step
        run (\verb|step_send|, \verb|step_recv|,
        \verb|step_internal|), composed with the operational T6
        chain to demonstrate that the existential statement
        \verb|CD_B_solvable_with_recv Unicast correct| is non-
        vacuously satisfiable -- there is an actual adversary,
        an actual run, and an actual algorithm output that
        witness it.
  \item \verb|T6_Multihop|: a three-process two-hop variant of
        the T6 demo.  Five \verb|run_step| transitions cover a
        bhb chain crossing three processes ($p_b$ sends to $p_a$,
        $p_a$ relays to $p_c$, $p_c$'s target internal event);
        \verb|multi_bhb_chain| proves the four-edge bhb path,
        \verb|T6_multihop_demo| and \verb|T6_multihop_witnessed|
        mirror the 1-message demo at the bigger scale.
  \item \verb|T6_With_Byzantine|: a T6 demo with a Byzantine
        bystander.  Two correct processes plus a Byzantine $p_c$
        performing an arbitrary local internal event (exercises
        \verb|step_byzantine|, unused in the other demos);
        \verb|byzantine_event_not_on_bhb_chain_*| proves the
        Byzantine's event is excluded from every bhb chain by
        \verb|bhb_proc_of_endpoints|; \verb|T6_with_byzantine_demo|
        and \verb|T6_with_byzantine_witnessed| demonstrate T6's
        robustness to live Byzantine activity.
  \item \verb|Causal_Scheduler|: scheduler-level refinement of
        \verb|run_step| that operationally realises BCB's
        causal-order delivery primitive (paper Section 4.3).
        Adds two side conditions to \verb|run_step|: freshness at
        \verb|step_send| (correct-to-correct triples are sent
        at most once) and a Byzantine-happened-before predecessor
        causal precondition at \verb|step_recv|.  Proves a joint
        invariant (send and receive uniqueness, buffer count,
        delivered-drained, receive-implies-send), the
        Byzantine-happened-before version of BCB causal-order
        (\verb|causal_run_satisfies_bhb_causal_order|), and
        end-to-end composition into operational T7
        (\verb|fair_drained_causal_run_solves_CD_B_broadcast|),
        closing the last operational hypothesis on the T7 chain.
  \item \verb|CO|: paper Section 5.2, Theorems 17 and 18.  The
        causal-ordering problem CO as a strict restriction of CD
        to receive-event targets; the forward reduction CD
        $\implies$ CO; T18 false-negative and false-negative-or-
        false-positive constructions; the impossibility of CO
        under all three communication modes; the interreducibility
        statement under Byzantine premises.
  \item \verb|CD_with_Crypto|: paper Section 4.4 + 4.5, Theorems
        9 through 14.  The six cryptography theorems, all as
        corollaries of the corresponding non-crypto results plus
        T14 as the new multicast-with-crypto-possibility case.
\end{enumerate}

\newpage

% include generated text of all theories
\input{session}

\nocite{*}
\bibliographystyle{abbrv}
\bibliography{root}

\end{document}
